***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-10-12 Schloss_Cluster_PeerJ_2015.Rmd
Incease scipen option
Entered some references

## To Do:
* references
* revise

@@

options(scipen=999)

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, known as operational taxonomic units (OTUs), are commonly used to compare the biodiversity contained within and between different samples. Such comparisons have enabled researchers to characterize the microbiota associated with the human body (REF), soil (REF), aquatic ecosystems (REF), and numerous other environments. Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within packages such as mothur and QIIME and stand-alone tools (REFS), it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping or closed reference clustering. This approach involves comparing sequences to a curated database and then clustering sequences together that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference sequences do not reflect the composition of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Because of this, defining OTUs in the closed-reference approach is complicated because two sequences might be 97% similar to the same reference sequence; however, they may only be 94% similar to each other. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level in the Linnean taxonomic hierarchy. Ultimately, this approach is an example of supervised classification.

The second approach has been referred to as *de novo* clustering. In this method, the distance between sequences is used to cluster sequence rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures then these problems can be overcome (REF). As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods. Previous comparisons of hierarchical and heuristic methods have shown that the average neighbor method, a hierarchical method outperforms the other *de novo* clustering method. An early analysis using 16S rRNA gene sequence data generated using the 454 sequencing platform found that if these methods were applied to the same dataset using a subset of the sequences then a different number of OTUs were observed compared to clustering the full dataset and rarefying to the same number of sequences. Because the initial subsampling of the dataset was not repeated in the earlier analysis, it is unclear whether the result was product of the method or sampling. One explanation could be that the clusters are getting better with additional sampling. Another observed problem with *de novo* clustering is that because it is necessary to break ties when assigning sequences to OTUs, clustering the same data multiple times may result in different clustering assignments. Whether the differences in assignments is meaningful is unclear; however the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU (ref). For this reason, *de novo* clustering has been preferred across the field. In contrast to closed-reference clustering, *de novo* clustering is an example of unsupervised clustering.

The third approach is a hybrid of the closed-reference and *de novo* approaches. This approach has been called open-reference clustering. This method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. This method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions generated by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels (REF). For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor method. Those sequences that did not classify to a known family would also be clustered using the average neighbor method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group (e.g. each family) is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

@@

>>>>

## To Do:
* references
* revise
* add sensitivity/specificity of closed-reference searching

@@

options(scipen=99999)

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, commonly known as operational taxonomic units, are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing]. Such comparisons have enabled researchers to characterize the microbiota associated with the human body [@human2012structure], soil [@shade2013streptomycin], aquatic ecosystems [@gilbert2012defining], and numerous other environments. Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@caporaso2010qiime], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping [@schloss2011assessing] or closed reference clustering [@navas2013advancing]. This approach involves comparing sequences to a curated database and then clustering sequences together that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference sequences do not reflect the composition of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Because of this, defining OTUs in the closed-reference approach is complicated because two sequences might be 97% similar to the same reference sequence; however, they may only be 94% similar to each other. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing]. Ultimately, this approach is an example of supervised classification.

The second approach has been referred to as distance-based [@schloss2011assessing] or *de novo* clustering [@navas2013advancing]. In this method, the distance between sequences is used to cluster sequence rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures then these problems can be overcome [@kozich2013development]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593]. Previous comparisons of hierarchical and heuristic methods have shown that the average neighbor method, a hierarchical method, outperformed the other *de novo* clustering methods. An early analysis using 16S rRNA gene sequence data generated using the 454 sequencing platform found that if these methods were applied to the same dataset using a subset of the sequences then a different number of OTUs were observed compared to clustering the full dataset and rarefying to the same number of sequences. Because the initial subsampling of the dataset was not repeated in the earlier analysis, it is unclear whether the result was product of the method or sampling. One explanation could be that the clusters are getting better with additional sampling. Another observed problem with *de novo* clustering is that because it is necessary to break ties when assigning sequences to OTUs, clustering the same data multiple times may result in different clustering assignments. Whether the differences in assignments is meaningful is unclear; however the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@schloss2011assessing]. For this reason, *de novo* clustering has been preferred across the field.

The third approach is a hybrid of the closed-reference and *de novo* approaches. This approach has been called open-reference clustering. This method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference [@navas2013advancing]. This method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions generated by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels [@schloss2011assessing]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor method. Those sequences that did not classify to a known family would also be clustered using the average neighbor method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group (e.g. each family) is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

@@

\bibliography
